---
title: "DESeq2_tonsa_multigen"
author: "Lauren Santanen"
date: "2025-10-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/users/l/s/lsantane/projects/eco_genomics_2025/transcriptomics/mydata")
```


```{r}
## Import the libraries that we're likely to need in this session

library(DESeq2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(ggpubr)
library(vsn)  
library("pheatmap")
library("vsn")

####################################################

### Import our data

####################################################

 
# Import the counts matrix
countsTable <- read.table("counts_matrix.txt", header=TRUE, row.names=1)
head(countsTable)
```


```{r}
dim(countsTable)
```

```{r}
countsTableRound <- round(countsTable) # bc DESeq2 doesn't like decimals (and Salmon outputs data with decimals)
head(countsTableRound)
```

```{r}
#import the sample description table
conds <- read.delim("metadata.txt", header=TRUE, stringsAsFactors = TRUE, row.names=1)
head(conds)
```
Explore data distributions

```{r}

# Let's see how many reads we have from each sample
colSums(countsTableRound)
```


```{r}
mean(colSums(countsTableRound))
```

Note that we are below the general rule of thumb recommendation of 20M reads per sample. We're at about 12M reads per sample

```{r}
barplot(colSums(countsTableRound), names.arg=colnames(countsTableRound),cex.names=0.5, las=3,ylim=c(0,21000000)) #change dims
abline(h=mean(colSums(countsTableRound)), col="blue", lwd=2)

```


```{r}
# the average number of counts per gene
rowSums(countsTableRound)
```

```{r}
mean(rowSums(countsTableRound)) # 

```

Despite lots of zeros across the 119,800 transcripts, we still have an average of 2500 reads mapping per transcript.

```{r}
median(rowSums(countsTableRound)) #
```


```{r}
apply(countsTableRound,2,mean) # 2 in the apply function does the action across columns
apply(countsTableRound,1,mean) # 1 in the apply function does the action across rows
```


```{r}
hist(apply(countsTableRound,1,mean),xlim=c(0,15000), ylim=c(0,100),breaks=10000)
```
#Now that we've explored the data, let's run DESeq2

```{r}

#Corrects the column names to match between the metadata table and the counts matrix
colnames(countsTableRound) <- substr(colnames(countsTableRound), start = 1, stop = 4)

####################################################

### Start working with DESeq2!

####################################################

#### Create a DESeq object and define the experimental design here with the tilda

dds <- DESeqDataSetFromMatrix(countData = countsTableRound, colData=conds, 
                              design= ~ generation + line + generation:line)

dim(dds)

# Filter out genes with too few reads - remove all genes with counts < 15 in more than 75% of 24 samples, so 18)
## suggested by WGCNA on RNAseq FAQ

dds <- dds[rowSums(counts(dds) >= 15) >= 18,]
nrow(dds) 
# How many transcripts have at least 15 reads (a.k.a counts) in 75% of the samples

# Run the DESeq model to test for differential gene expression
dds <- DESeq(dds)

# List the results you've generated
resultsNames(dds)
# Copy the results names 
```
Continuing on 10/16/25

```{r}
# The goal of transformation "is to remove the dependence of the variance on the mean, particularly the high variance of the logarithm of count data when the mean is low."

# this gives log2(n + 1)
ntd <- normTransform(dds)
meanSdPlot(assay(ntd))

# Variance stabilizing transformation
vsd <- vst(dds, blind=FALSE)
meanSdPlot(assay(vsd))

```

Use the variance stabilized data to make plots
Look for outliers

```{r}
sampleDists <- dist(t(assay(vsd)))

library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$line, vsd$generation, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

# Note any outliers
```
Make a cluster tree to look for outlier samples 

```{r}
sampleTree <- hclust(dist(sampleDists), method="average")
# plot
plot(sampleTree, main="Sample clustering to detect outliers", sub="", xlab="",cex.lab=1.5, cex.axis=1.5, cex.main=2)
```

On to PCA...

```{r}
# first transform the data for plotting using variance stabilization
vsd <- vst(dds, blind=FALSE)

pcaData <- plotPCA(vsd, intgroup=c("line","generation"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData,"percentVar"))

ggplot(pcaData, aes(PC1, PC2, color=line, shape=generation)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

Optimizing data visualization
Added in additional detail
Points are bigger, added shapes and indicated what they are, treatments are blue vs red

```{r}
ggplot(pcaData, aes(PC1, PC2)) +
  geom_point(size=5, stroke = 1, aes(fill=line, shape=generation, alpha = generation)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  scale_shape_manual(values=c(21,22,23,24), labels = c("G1", "G2","G3", "G4"))+
  #scale_shape_manual(values=c(21,22,23,24) -> circle, square, diamond, triangle
  scale_fill_manual(values=c('#6699CC',"#CC3333"), labels = c("Control", "Treatment"))+
  scale_alpha_manual(values = c(1.0, 0.8, 0.7, 0.3), guide = "none") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 3))+
  theme(text = element_text(size = 20)) +
  theme(legend.title = element_blank())
```
Differential gene expression testing
Making groups to test specific contrasts

```{r}
dds$line <- relevel(dds$line, ref = "control")
dds$generation <- relevel(dds$generation, ref = "G1")


dds$group <- factor(paste0(dds$line, dds$generation))
design(dds) <- ~ group
dds <- DESeq(dds)
resultsNames(dds)

```

Pull out results of interest using control generation 1 and treatment generation 1 with a sig value of .05

Can change the order that the contrast is in (control and treatment) and the log2FoldChange will switch signs

```{r}
resG1_CvT <- results(dds, contrast=c("group","controlG1","treatmentG1"), alpha = 0.05)

summary(resG1_CvT)
```
Reorder by significance

```{r}
resG1_CvT <- resG1_CvT[order(resG1_CvT$padj),]
head(resG1_CvT)
```

Plot individual genes

```{r}
# Counts of specific top interaction gene! (important validatition that the normalization, model is working)
d <-plotCounts(dds, gene="TRINITY_DN93941_c0_g3_i1", intgroup = (c("line","generation")), returnData=TRUE)
d

p <-ggplot(d, aes(x=line, y=count, color=line, shape=generation)) + 
  theme_minimal() + theme(text = element_text(size=20), panel.grid.major=element_line(colour="grey"))
p <- p + geom_point(position=position_jitter(w=0.2,h=0), size=3)

p <- p + stat_summary(fun = mean, geom = "point", size=5, alpha=0.7) 
p
```

