---
title: "HW1_PCA_Admix"
author: "Lauren Santanen"
date: "2025-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_knit$set(root.dir="~/projects/eco_genomics_2025/population_genomics/myresults/ANGSD/PCA_ADMIX/")
```

Load libraries for plotting

```{r}
library(ggplot2)
library(ggpubr)
```
First, let's estimate our PCA!

```{r}
#K=2
COV2 <- as.matrix(read.table("RSBS_poly_K2.cov"))
PCA2 <- eigen(COV2)

#K=3
COV3 <- as.matrix(read.table("RSBS_poly_K3.cov"))
PCA3 <- eigen(COV3)

#K=4
COV4 <- as.matrix(read.table("RSBS_poly_K4.cov"))
PCA4 <- eigen(COV4)

#K=5
COV5 <- as.matrix(read.table("RSBS_poly_K5.cov"))
PCA5 <- eigen(COV5)

```


How much variance does the first eigenvalue explain?
And screeplot of each K value

K=2

```{r}
#K=2
varK2 <- round(PCA2$values/sum(PCA2$values),3)
varK2[1:3]

barplot(varK2, 
        xlab="Eigenvalues of the PCA K=2", 
        ylab="Proportion of variance explained K=2")

```

K=3
```{r}
#K=3
varK3 <- round(PCA3$values/sum(PCA3$values),4)
varK3[1:4]

barplot(varK3, 
        xlab="Eigenvalues of the PCA K=3", 
        ylab="Proportion of variance explained K=3")
```

K=4
```{r}
#K=4
varK4 <- round(PCA4$values/sum(PCA4$values),5)
varK4[1:5]

barplot(varK4, 
        xlab="Eigenvalues of the PCA K=4", 
        ylab="Proportion of variance explained K=4")
```

K=5
```{r}
#K=5
varK5 <- round(PCA5$values/sum(PCA5$values),6)
varK5[1:6]

barplot(varK5, 
        xlab="Eigenvalues of the PCA K=5", 
        ylab="Proportion of variance explained K=5")
```

##STOPPED HERE ON 10/3##

Data wrangling that takes in the names of our samples and makes them into a tidy looking format that we can use for plotting

```{r}
names <- read.table("RSBS_bam.list")
names <- unlist(strsplit(basename(as.character(names[,1])), split = ".sorted.rmdup.bam"))
split = strsplit(names, "_")
pops <- data.frame(names[1:113], do.call(rbind, split[1:113]))
names(pops) = c("Ind", "Pop", "Row", "Col")

data=as.data.frame(PCA$vectors)
data=data[,c(1:5)]
data= cbind(data, pops)
```
Let's plot the PCA
Make graphs for K=2,3,4,5

```{r}
#K=2
ggscatter(data, x = "V1", y = "V2",
          color = "Pop",
          mean.point = TRUE,
          star.plot = TRUE,
          main="Red spruce-Black spruce genetic PCA") +
  theme_bw(base_size = 13, base_family = "Times") +
  labs(x = paste0("PC1: (",varK2[1]*100,"%)"), y = paste0("PC2: (",varK2[2]*100,"%)"))


#K=3
ggscatter(data, x = "V1", y = "V3",
          color = "Pop",
          mean.point = TRUE,
          star.plot = TRUE,
          main="Red spruce-Black spruce genetic PCA") +
  theme_bw(base_size = 13, base_family = "Times") +
  labs(x = paste0("PC1: (",varK3[1]*100,"%)"), y = paste0("PC3: (",varK3[3]*100,"%)"))

#K=4
ggscatter(data, x = "V1", y = "V4",
          color = "Pop",
          mean.point = TRUE,
          star.plot = TRUE,
          main="Red spruce-Black spruce genetic PCA") +
  theme_bw(base_size = 13, base_family = "Times") +
  labs(x = paste0("PC1: (",varK4[1]*100,"%)"), y = paste0("PC4: (",varK4[4]*100,"%)"))

#K=5
ggscatter(data, x = "V1", y = "V5",
          color = "Pop",
          mean.point = TRUE,
          star.plot = TRUE,
          main="Red spruce-Black spruce genetic PCA") +
  theme_bw(base_size = 13, base_family = "Times") +
  labs(x = paste0("PC1: (",varK5[1]*100,"%)"), y = paste0("PC5: (",varK5[5]*100,"%)"))
```
What about the admixture analysis?

```{r}
#K=2
q2 <- read.table("RSBS_poly_K2.admix.2.Q", sep=" ", header=F)
K2=dim(q2)[2] #Find the level of K modeled
## rank order the samples according to population code
ord2<-order(pops[,2])


#K=3
q3 <- read.table("RSBS_poly_K3.admix.3.Q", sep=" ", header=F)
K3=dim(q3)[2] #Find the level of K modeled
## rank order the samples according to population code
ord3<-order(pops[,2])

#K=4
q4 <- read.table("RSBS_poly_K4.admix.4.Q", sep=" ", header=F)
K4=dim(q4)[2] #Find the level of K modeled
## rank order the samples according to population code
ord4<-order(pops[,2])

#K=5
q5 <- read.table("RSBS_poly_K5.admix.5.Q", sep=" ", header=F)
K5=dim(q5)[2] #Find the level of K modeled
## rank order the samples according to population code
ord5<-order(pops[,2])

```

Plot the admixture coefficients 

```{r}
#K=2
barplot(t(q2)[,ord2],
        #col=cols[1:K],
        col=c("black","red"),
        space=0,border=NA,
        xlab="Populations",ylab="Admixture proportions",
        main=paste0("Red Spruce-Black Spruce Admixture: K=",K2))
text(tapply(1:nrow(pops),pops[ord2,2],mean),-0.05,unique(pops[ord2,2]),xpd=T,cex=0.75,srt=45)
abline(v=cumsum(sapply(unique(pops[ord2,2]),function(x){sum(pops[ord2,2]==x)})),col=1,lwd=1.2)

#K=3
barplot(t(q3)[,ord3],
        #col=cols[1:K],
        col=c("black","red"),
        space=0,border=NA,
        xlab="Populations",ylab="Admixture proportions",
        main=paste0("Red Spruce-Black Spruce Admixture: K=",K3))
text(tapply(1:nrow(pops),pops[ord3,2],mean),-0.05,unique(pops[ord3,2]),xpd=T,cex=0.75,srt=45)
abline(v=cumsum(sapply(unique(pops[ord3,2]),function(x){sum(pops[ord3,2]==x)})),col=1,lwd=1.2)

#K=4
barplot(t(q4)[,ord4],
        #col=cols[1:K],
        col=c("black","red"),
        space=0,border=NA,
        xlab="Populations",ylab="Admixture proportions",
        main=paste0("Red Spruce-Black Spruce Admixture: K=",K4))
text(tapply(1:nrow(pops),pops[ord4,2],mean),-0.05,unique(pops[ord4,2]),xpd=T,cex=0.75,srt=45)
abline(v=cumsum(sapply(unique(pops[ord4,2]),function(x){sum(pops[ord4,2]==x)})),col=1,lwd=1.2)

#K=5
barplot(t(q5)[,ord5],
        #col=cols[1:K],
        col=c("black","red"),
        space=0,border=NA,
        xlab="Populations",ylab="Admixture proportions",
        main=paste0("Red Spruce-Black Spruce Admixture: K=",K5))
text(tapply(1:nrow(pops),pops[ord5,2],mean),-0.05,unique(pops[ord5,2]),xpd=T,cex=0.75,srt=45)
abline(v=cumsum(sapply(unique(pops[ord5,2]),function(x){sum(pops[ord5,2]==x)})),col=1,lwd=1.2)
```


