---
title: "HW1_PCA_Admix"
author: "Lauren Santanen"
date: "2025-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_knit$set(root.dir="~/projects/eco_genomics_2025/population_genomics/myresults/ANGSD/PCA_ADMIX/")
```

Load libraries for plotting

```{r}
library(ggplot2)
library(ggpubr)
```
First, let's estimate our PCA!

```{r}
#K=2
COV <- as.matrix(read.table("RSBS_poly_K2.cov"))
PCA <- eigen(COV)

#K=3
COV <- as.matrix(read.table("RSBS_poly_K3.cov"))
PCA <- eigen(COV)

#K=4
COV <- as.matrix(read.table("RSBS_poly_K4.cov"))
PCA <- eigen(COV)

#K=5
COV <- as.matrix(read.table("RSBS_poly_K5.cov"))
PCA <- eigen(COV)

```


How much variance does the first eigenvalue explain?
And screeplot of each K value

K=2

```{r}
#K=2
var <- round(PCA$values/sum(PCA$values),3)
var[1:3]

barplot(var, 
        xlab="Eigenvalues of the PCA K=2", 
        ylab="Proportion of variance explained K=2")
```

K=3
```{r}
#K=3
var <- round(PCA$values/sum(PCA$values),4)
var[1:4]

barplot(var, 
        xlab="Eigenvalues of the PCA K=3", 
        ylab="Proportion of variance explained K=3")
```

K=4
```{r}
#K=4
var <- round(PCA$values/sum(PCA$values),5)
var[1:5]

barplot(var, 
        xlab="Eigenvalues of the PCA K=4", 
        ylab="Proportion of variance explained K=4")
```

K=5
```{r}
#K=5
var <- round(PCA$values/sum(PCA$values),6)
var[1:6]

barplot(var, 
        xlab="Eigenvalues of the PCA K=5", 
        ylab="Proportion of variance explained K=5")
```

##STOPPED HERE ON 10/3##

Data wrangling that takes in the names of our samples and makes them into a tidy looking format that we can use for plotting

```{r}
names <- read.table("RSBS_bam.list")
names <- unlist(strsplit(basename(as.character(names[,1])), split = ".sorted.rmdup.bam"))
split = strsplit(names, "_")
pops <- data.frame(names[1:113], do.call(rbind, split[1:113]))
names(pops) = c("Ind", "Pop", "Row", "Col")

data=as.data.frame(PCA$vectors)
data=data[,c(1:3)]
data= cbind(data, pops)
```
Let's plot the PCA

```{r}
ggscatter(data, x = "V1", y = "V2",
          color = "Pop",
          mean.point = TRUE,
          star.plot = TRUE,
          main="Red spruce-Black spruce genetic PCA") +
  theme_bw(base_size = 13, base_family = "Times") +
  labs(x = paste0("PC1: (",var[1]*100,"%)"), y = paste0("PC2: (",var[2]*100,"%)"))
```
What about the admixture analysis?

```{r}
q <- read.table("RSBS_poly_K2.admix.2.Q", sep=" ", header=F)

K=dim(q)[2] #Find the level of K modeled

## rank order the samples according to population code
ord<-order(pops[,2])
```

Plot the admixture coefficients 

```{r}
barplot(t(q)[,ord],
        #col=cols[1:K],
        col=c("black","red"),
        space=0,border=NA,
        xlab="Populations",ylab="Admixture proportions",
        main=paste0("Red Spruce-Black Spruce Admixture: K=",K))
text(tapply(1:nrow(pops),pops[ord,2],mean),-0.05,unique(pops[ord,2]),xpd=T,cex=0.75,srt=45)
abline(v=cumsum(sapply(unique(pops[ord,2]),function(x){sum(pops[ord,2]==x)})),col=1,lwd=1.2)
```


